```{r}
library(tidyverse)
```

Here's an example of how to calculate z-scores using NCAA volleyball teams. We'll use match data for the 2023 season.

```{r}
matches23 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/main/data/ncaa_womens_volleyball_matchstats_2023.csv")

matches24 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/main/data/ncaa_womens_volleyball_matchstats_2024.csv")
```

Get Maryland's Aces

```{r}
matches23 |> 
  filter(team == "Maryland Terrapins, Terps") |> 
  group_by(result) |>
  summarize(
    avg_aces = mean(aces)
  )

matches24 |> 
  filter(team == "Maryland Terrapins, Terps") |> 
  group_by(result) |>
  summarize(
    avg_aces = mean(aces)
  )
```


T-test for significance
h0: no difference between the two years
ha: there is a difference between the two years

```{r}
umd_aces23 <- matches23 |> filter(team == "Maryland Terrapins, Terps") |> mutate(season=2023) #|>filter(date < '2023-09-12')
  
umd_aces24 <- matches24 |> filter(team == "Maryland Terrapins, Terps") |> mutate(season=2024)

result <- t.test(umd_aces23$aces, umd_aces24$aces, alternative = "two.sided", var.equal = TRUE)

print(result)
```

ANOVA test

```{r}
combined <- bind_rows(umd_aces23,umd_aces24)

combined$result <- factor(combined$result)
combined$season <- factor(combined$season)

anova_result <- aov(aces ~ result * season, data = combined)

summary(anova_result)
```







### Isolate the columns we might want to use for our metrics

```{r}
teamquality <- matches |> 
  select(team, kills, s, errors, total_attacks, assists, s_err, digs, ret_att, r_err, b_err)
```


### Calculate averages

```{r}
teamtotals <- teamquality |> 
  group_by(team) |> 
  summarise(
    avg_kills = mean(kills),
    avg_errors = mean(errors),
    avg_attacks = mean(total_attacks),
    avg_assists = mean(assists),
    avg_s_err = mean(s_err),
    avg_digs = mean(digs),
    avg_ret_att = mean(ret_att),
    avg_r_err = mean(r_err),
    avg_b_err = mean(b_err)
  )
```

### Create Z-Scores

```{r}
teamzscore <- teamtotals |> 
  mutate(
    kills_zscore = as.numeric(scale(avg_kills, center = TRUE, scale = TRUE)),
    errors_zscore = as.numeric(scale(avg_errors, center = TRUE, scale = TRUE)) * -1,
    s_err_zscore = as.numeric(scale(avg_s_err, center = TRUE, scale = TRUE)) * -1,
    digs_zscore = as.numeric(scale(avg_digs, center = TRUE, scale = TRUE)),
    TotalZscore = kills_zscore + errors_zscore + s_err_zscore + digs_zscore
  )
```

### Look at the results

```{r}
teamzscore |> select(team, TotalZscore, kills_zscore, errors_zscore, s_err_zscore, digs_zscore) |>  arrange(desc(TotalZscore))
```





***Z SCORES ANALYSIS***
```{r}
library(tidyverse)
matches24<-read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/main/data/ncaa_womens_volleyball_matchstats_2024.csv")
```

Set up a code block to create team averages and z-scores the way the tutorial did, but you must change the metrics you use to calculate the scores so that you're using a different measure than the tutorial did.
```{r}
team_quality_vb<-matches24|>
  select(team,kills,hit_pct,total_attacks,aces,digs,block_solos)
```

```{r}
team_totals_vb <- team_quality_vb |>
  group_by(team) |> 
  summarise(
    killavg = mean(kills),
    attacksavg = mean(total_attacks),
    hitpctavg = mean(hit_pct),
    acesavg = mean(aces),
    digsavg = mean(digs),
    blocks_savg = mean(block_solos)
  )
```

```{r}
team_z_score_vb <- team_totals_vb |>
  mutate(
  killZscore = as.numeric(scale(killavg, center = TRUE, scale = TRUE)),
  attackZscore = as.numeric(scale(attacksavg, center = TRUE, scale = TRUE)),
  hitpctZscore = as.numeric(scale(hitpctavg, center = TRUE, scale = TRUE)),
  acesZscore = as.numeric(scale(acesavg, center = TRUE, scale = TRUE)),
  digsZscore = as.numeric(scale(digsavg, center = TRUE, scale = TRUE)),
  blocksZscore = as.numeric(scale(blocks_savg, center = TRUE, scale = TRUE)),
  total_z_score = killZscore + attackZscore + hitpctZscore + acesZscore + digsZscore + blocksZscore)

team_z_score_vb |> arrange(desc(total_z_score))
```


When you do this, who are the highest-performing teams in the NCAA? Where does Maryland rank on the Z-score scale this season and what teams are similar to it? Use code blocks to generate those lists/answers, and be sure to add descriptions in text, too. What are the best story ideas here?

```{r}
team_z_score_vb |> filter(team == "Maryland")
```
Unfortunately, there was no filter for conference, so this is the best I can do for filtering, and manually scrolling/sorting. Maryland's total Z score is 2.36, which is above average and in any normal distribution would be , but actually ranks 95th in the nation when it comes to this data. There has been a wide distribution of these statistics across the NCAA, so this explains why they are ranked lower than a statistician would expect. They are doing very well in the aces category, which we discussed earlier in the semester, ranking fourth in the country with their aces z score of 2.3. Aside from hit pct as a positive, the other statistics don't stand out much for the terps - three of them are in the negatives for z score. 

```{r}
teams <- c("Maryland", "Virginia", "Providence", "Georgia Tech")

filtered_teams <- team_z_score_vb %>%
  filter(team %in% teams)|>
  arrange(desc(total_z_score))
print(filtered_teams)
```
What has caused Providence to be less successful than the other teams in a similar position as them? The team is barely breaking .500 with a 7-6 record, compared to the other three teams measured who are all around .700 or above. They have also recently lost their first two Big East matches in straight sets. 

On the flip side, Georgia Tech is a top 15 team in the country - why are they so far down on this list? Although they are below average in multiple facets of the game, their hit percent and aces are carrying them to victories. Hit percentage is a reflection of a very disciplined team that makes very little errors - I wonder what the coaches do to harp upon this, does it come from who they recruit or how they coach? Is it a mindset thing or a skill thing for Georgia Tech? 

The same is true for Maryland - they are very strong in hit percent and aces, so why aren't we ranked? Does this show our potential? Have other winning teams in the past been carried by these stats, too?


```{r}
team_z_score_vb |> filter(team=="SIUE")
```

*Short Story on Findings*

Can a bad team fake being good? When it comes down to the hardcore stats in volleyball - hits, aces, digs, blocks, kills, and attack, Southern Illinois University Edwardsville (SIUE) finds itself sitting far beyond average compared to the rest of the NCAA. In fact, when combining its statistical measures and measuring them against the country, it is in the 99th percentile. However, the team shows a 5-8 record through a significant chunk of their season. What is going wrong with this team that is expected to be good?

SIUE particularly sets itself apart with its average attacks/game (148) and solo blocks/set (2.85). To compare, it the blows top NCAA team Pitt out of the conversation (89 and 1.25 respectively) with these numbers. Freshman Taya Nordmann has been leading the charge for the team, with 1.27 blocks per set. Every other player but one on the team contributes half as many blocks per set - meaning she carries most of the load. The defense is anchored by a freshman, and it is clear that the leadership isn't quite there for the team. As they continue to develop, they may look to the freshman for inspiration, but for now, their lack of experience on defense has cost them many games. 

In terms of attacking, the team is led by three upperclassmen who have carried the load for the past 3+ years. However, after their leading attackers, there is a significant dropoff in production. Depth can be an issue, especially for a smaller school in a small conference. They have won at least 1 set in 11/13 games, but have failed to finish strong in many. The stars on their squad have boosted their statistics, but the rest of the team fails to back it up with depth and production from the rest of their team and their defense. 

### Dumbbell/Lollipop plots 10/17/24
```{r}
library(tidyverse)
library(ggalt)
```

```{r}
logs <- ncaa_womens_volleyball_matchstats_2023_1_
```

```{r}
big10 <- c("Nebraska Cornhuskers", "Iowa Hawkeyes", "Minnesota Golden Gophers", "Illinois Fighting Illini", "Northwestern Wildcats", "Wisconsin Badgers", "Indiana Hoosiers", "Purdue Boilermakers", "Ohio St. Buckeyes", "Michigan Wolverines", "Michigan St. Spartans", "Penn St. Nittany Lions", "Rutgers Scarlet Knights", "Maryland Terrapins, Terps")

errors <- logs |>
  filter(team %in% big10) |> 
  group_by(team) |> 
  summarise(
    total_errors = sum(errors), 
    opp_errors = sum(defensive_errors))
```

```{r}
ggplot() + 
  geom_dumbbell(
    data=errors, 
    aes(y=reorder(team, opp_errors), x=opp_errors, xend=total_errors),
    linewidth = 1,
    color = "grey",
    colour_x = "green",
    colour_xend = "red") + 
  theme_minimal()
```



```{r}
md <- errors |> filter(team == "Maryland Terrapins, Terps")
```

```{r}
ggplot() + 
  geom_lollipop(
    data=errors, 
    aes(y=reorder(team, opp_errors), x=opp_errors), 
    horizontal = TRUE
    ) + 
  geom_lollipop(
    data=md,
    aes(y=team, x=opp_errors),
    horizontal = TRUE,
    color = "red"
  ) + 
  theme_minimal() + 
  labs(title = "Maryland middle of the pack among Big Ten teams", y="team")
```

