First Presentation - Data Backgrounding

Guiding Question:
What early career stats help predict a player's upward or downward career trajectory over time? 

There are many subquestions that can be asked within this based on numerous factors. Is there a 'year' that players must be producing by until teams give up on them? Is instant success or gradual improvement better? What about their role/usage? 


I plan on using data from the 2003 season to the present day (had to include LeBron's rookie year), and collecting it from basketballreference. This website has pretty much everything I need, including advanced stats/metrics from every player. I probably won't need assistance in getting it, I will just figure out the best method to get all of this data into one place - ie. if one season per file is best, or if I should load them all into one big file. 

Basketballreference sometimes has some funky formatting, for example if a player played for multiple teams in a season, they have separate stats and then a total stat - there is a team classified as '2TM' that I will need to work through. They also have headers every 20 or so players, which will just be annoying to remove for every season. Advanced stats and usage stuff that I may use are also on a separate tab from the 'basic' stats, so combining those may be a struggle, too. Calculating someone's 'years pro' may be tough, too - I haven't been able to find it on the website.


*Some Harvard students did a similar study about player development in the NBA back in 2015 (link: https://harvardsportsanalysis.org/2015/05/player-progression-in-the-nba/). 

*A man did try to predict NBA success based on the first few years and used a number of different measures/presentations (I really liked his decision trees), it gave me a few ideas but also had some differences.  https://www.kaggle.com/code/tombutton/predicting-nba-success

MIT's Sloan school which runs a lot of Sports Analytics stuff also published a paper about predicting success for players using both their pre-NBA and NBA data using a lot of advanced statistics. They focused on something called relevance-based prediction. (https://mitsloan.mit.edu/shared/ods/documents?PublicationDocumentID=10079)

All of these pieces are similar to what I'm doing, but aren't exactly the same. It does provide me with some preliminary info on what may be good/bad but I also think that I have a bit of my own path to chart with this. 

I have tried a number of different ways to get NBA data from each season. I'm wondering if there are any preloaded packages that I may be able to use. It's especially important for me to have the 'years pro' section for each player. I have tried manually downloading each CSV file for each NBA season (I did so for about the most recent 5 seasons, but realized that it was going to be a bit tedious), and have also tried to use hoopR and nbastatR, but may need assistance in how to obtain data from the past 21 NBA seasons. Below are some attempts, but I'm not sure where it gets me. 

```{r}
install.packages("hoopR")
install.packages("tictoc")
```

```{r}
tictoc::tic()
progressr::with_progress({
  nba_pbp <- hoopR::load_nba_pbp(season = 2024)
})
tictoc::toc()
```

Load NBA 2024 season stats
```{r}
nba24 <- read_excel("~/Downloads/nba24.xlsx", 
+     sheet = "nba24")
```


```{r}
devtools::install_github("abresler/nbastatR", force = TRUE)
```
```{r}
library(nbastatR)
```

