

Essentially, I will be checking to see if players really thought that they had "home court advantage" even though everyone was playing in the same spot - would players feel more confident in their home jerseys or with the fact that they get a certain bench on one side of the floor?

*NULL HYPOTHESIS*: There is no significant difference between winning percentage and whether a team was listed as "home" or "away" during the NBA bubble.
*ALTERNATIVE HYPOTHESIS* There is a significant difference between winning percentage and home/away status during the NBA bubble.

```{r}
library(tidyverse)

logs <- read.csv("https://dwillis.github.io/sports-data-files/nbabubble.csv")

head(logs)
```


Let's get every game without fans and find the average FGPCT just to see (*this was another test, not related to this*)
```{r}
logs |> 
  filter(COVID == "Without Fans") |>
  summarize(
    avg_fgp = mean(TeamFGPCT)
  )


logs |> 
  filter(COVID == "With Fans") |>
  summarize(
    avg_fgp = mean(TeamFGPCT)
  )

```

Need winning percentage.
```{r}
logs |> 
  filter(COVID == "Without Fans") |>
  filter(W_L == "W") |>
  summarize(
    wins = n()
  ) |>
```


```{r}
logs |> 
  filter(COVID == "Without Fans") |>
  filter(W_L == "L") |>
  summarize(
    losses = n()
  )
```
Almost there!



Now, let's test them.





This here below shows difference of FGPct whether games were "home" or "away" in the bubble (which really wasn't a thing, I just wanted to see if there actually was any difference).
```{r}

fgpct_home <- logs |> filter(COVID == "Without Fans", is.na(HomeAway)) |> mutate(location="home")
fgpct_away <- logs |> filter(COVID == "Without Fans", HomeAway == "@") |> mutate(location="away")

result <- t.test(fgpct_home$TeamFGPCT, fgpct_away$TeamFGPCT, alternative = "two.sided", var.equal = TRUE)

print(result)


```

The p-value is much higher than 0.05, meaning we fail to reject the null hypothesis, meaning there was no
