---
output:
  html_document: default
  pdf_document: default
---

**Which NBA teams are best at developing their players?**

Sam Benning

JOUR479X Final Presentation

This presentation takes a bit of a turn on my first project, where I assessed which statistical measures are useful for predicting an NBA career. I continued to work with time-series-based data, looking at player development. Overall, I wanted to analyze the system of drafting and improving players throughout the NBA, analyzing by different teams, positions, and players.

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(gt)
library(plotly)
library(cowplot)
library(gt)
```

I scraped data from basketball reference, three different data sets from years 2005-2025 - 'basic' stats, advanced stats, and each team's NBA draft history. They are all combined into one file that includes a separate observation for a player from every season they played, with each year as an entry. Their draft data (team, rd, pick, traded), is all included as well.

```{r}
all_data <- read.csv("https://raw.githubusercontent.com/sbenning7/jour479x_fall_2024/refs/heads/main/presentations/presentation2/nba_0525.csv")

head(all_data)
```

```{r}
#Compute Top Players with Draft Year
top_players <- all_data |> 
  group_by(First_Team, Player) |> 
  filter(years_pro < 3, G > 20, MP > 200) |> 
  summarize(
    per = mean(PER, na.rm = TRUE),
    draft_year = unique(Draft_Yr),
    pick = unique(Pick),
    .groups = "drop"
  ) |> 
  arrange(First_Team, desc(per)) |> 
  group_by(First_Team) |> 
  mutate(rank = row_number()) |> 
  filter(rank <= 3) |>  # Top 3 players per team
  summarize(
    top_players = paste0(Player, " (", round(per, 1), ", ", draft_year, " ", "Pick ", pick, ")",
      collapse = "; "
    ),
    .groups = "drop"
  )

#Calculate Team PER
teamdrafting <- all_data |> 
  group_by(First_Team, Player) |> 
  filter(years_pro < 3, G > 20, MP > 200) |> 
  summarize(
    per = mean(PER, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  group_by(First_Team) |> 
  summarize(
    teamper = mean(per),
    .groups = "drop"
  ) |> 
  arrange(desc(teamper)) |> 
  left_join(top_players, by = "First_Team")

#Format Styled Table
styled_table <- teamdrafting |> 
  select(Team = First_Team, `Team PER` = teamper, `Top Players` = top_players) |> 
  gt() |> 
  tab_header(
    title = "Good Drafting = Nothing?",
    subtitle = "The NBA best and worst teams over the past 20 years boast strong drafting abilities."
  ) |> 
  cols_align(align = "center", columns = everything()) |> 
  fmt_number(
    columns = `Team PER`,
    decimals = 2
  ) |> 
  cols_label(
    Team = "Team",
    `Team PER` = "Avg Player PER (First 3yrs of Career)",
    `Top Players` = "Top 3 Players (PER, Draft Info)"
  ) |> 
  tab_style(
    style = list(cell_fill(color = "#f5f5f5")),
    locations = cells_body(columns = `Top Players`)
  ) |> 
  tab_source_note(
    source_note = md("**By:** Sam Benning  |  **Source:** [Basketball Reference](https://www.basketball-reference.com/)")
  ) |> 
  tab_options(
    table.border.top.color = "gray",
    table.border.bottom.color = "gray",
    heading.align = "center",
    column_labels.font.weight = "bold"
  )

# Display the styled table
styled_table
```

This chart gives us a good general idea of who is drafting solid guys - taking just players early career PER in my opinion was the best way to show this. I wouldn't have predicted New Orleans at the top, but it makes sense given that they have drafted three true superstar players in recent history, which is very rare across the league. Right behind them is Houston, who doesn't boast any superstars at the top of their list, which shows some consistency.

## Team's Developmental Abilities

If we want to take a peek into long term development, we can look at a team's ability to develop players over time. After doing some thinking, I figured that you could gain a lot of information based upon someone's rookie PER compared to their peak PER instead of their full career PER (because some players PER tanks at the end of their career).

I took the three current championship-level teams, and three mediocre teams (these teams also happened to be some of the best and worst in terms of PER growth). I will look at where they drafted players, and then how strong their PER difference is. This shows us who is drafting "diamonds in the rough" -- who might have the ability to find players with potential, and then develop them. What's the secret to success, and failure in terms of player personnel?

```{r}
#selecting the teams and getting the data
#chose players who weren't traded on draft night
teams_pd <- all_data |> filter(First_Team %in% c("GSW", "DEN", "WAS", "BOS", "BRK", "DET")) %>% filter(Traded == 0) |> filter(G > 10) |> filter(years_with_team >1) |> group_by(Player, First_Team)|>
  summarize(
    pick = mean(Pick),
    games = sum(G),
    min = sum(MP),
    draftyr = mean(Draft_Yr), #make this an average so that every year of a player's career isn't shown
    lastyr = max(Year), #shows how long a player's career lasted when next to draft year
    rookie_PER = first(PER),
    peak_PER = max(PER, na.rm = TRUE),
    PER_growth = peak_PER - rookie_PER,
    )

# Define team colors for formatting
team_colors <- c(
  "GSW" = "yellow", 
  "DEN" = "blue3",
  "BOS" = "green4", 
  "WAS" = "#E03A3E",
  "BRK" = "black",  
  "DET" = "#C8102E" 
)

# Reorder the Team_Drafted variable for visual
teams_pd <- teams_pd %>%
  mutate(
    First_Team = factor(First_Team, levels = c("GSW", "DEN", "BOS", "WAS", "BRK", "DET"))
  )

# Create the plot
ggplot() + 
  geom_point(
    data = teams_pd, 
    aes(x = draftyr, y = pick, size = PER_growth, color = First_Team), 
    alpha = 0.7
  ) +
  facet_wrap(~First_Team) +
  geom_text_repel(data=teams_pd, 
    aes(x=draftyr, y=pick, label=Player), size = 1.25) +
  scale_y_reverse() +
  scale_color_manual(values = team_colors) +  # Use the custom color palette
  labs(
    title = "There's More Than One Way to do it",
    subtitle = "Both good and bad teams can develop players well - what matters is everything else.",
    x = "Draft Year",
    y = "Draft Pick",
    size = "PER Growth",
    color = "Team"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 12),  # Bold facet titles
    plot.title = element_text(size = 16, face = "bold"),  # Bold title
    legend.position = "right"
  )
```

The story that I see is that the Pistons and Wizards are actually decent at player development - their chart doesn't look much different from GSW/BOS in fact. I think it reflects poor roster management as a reason that they can't succeed. Off the paper, it is clear that these teams just don't do well externally.

The conclusion that I come to is that in order to build a top-level team, you need both good drafting and development internally, and roster additions in free agency/trades. I mean, that seems pretty logical to me.

This makes the Warriors franchise truly stand out to me though - they didn't dip much into other players when they had their dynasty (aside from Durant).

## Who is best at developing by position?

We will furhter explore if each team has a particular specialty when it comes to development, looking specifically at each position. This helps show us where a team's developmental success is actually coming from. Again, PER growth is our stat here.

```{r}
# Analyze Player Development by Position
development_by_position <- all_data |> 
  group_by(Team, Player, Pos) |> 
  arrange(Player, Year) |>  # Ensure data is ordered by year
  mutate(
    years_with_team = cumsum(Team == lag(Team, default = first(Team)))  # Increment tenure
  ) |> 
  filter(years_with_team > 1, G > 20) |>  # Only include players with more than 1 year with a team
  summarize(
    first_per = PER[Year == min(Year)],  # PER in the first year with the team
    peak_per = max(PER, na.rm = TRUE),  # Peak PER with the team
    improvement = peak_per - first_per,  # Improvement
    years_with_team = max(years_with_team),
    .groups = "drop")
```

```{r}
#Table to display best developers by position
team_pos_dev <- development_by_position |>
  group_by(Team, Pos) |> 
  summarize(
    avg_starting_per = round(mean(first_per, na.rm = TRUE), 2),
    avg_peak_per = round(mean(peak_per, na.rm = TRUE), 2),
    avg_improvement = round(mean(improvement, na.rm = TRUE), 2),
    player_count = n(),
    .groups = "drop"
  ) |> 
  arrange(desc(avg_improvement)) |> slice_head(n=10)

# Create the GT table
team_pos_dev |> 
  gt() |>
  cols_label(
    avg_starting_per = "Avg. PER entering team",
    avg_peak_per = "Avg. Peak PER while with team (after development)",
    avg_improvement = "Avg. Improvement",
    player_count = "# of Players"
  ) |>
  tab_header(
    title = "GSW Stars with Curry's Development",
    subtitle = "These 10 teams have excelled at helping players at certain positions improve their Player Efficiency Rating (PER) since 2005."
  ) |> 
  tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |>
  tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),  # Bold the Avg. Improvement column
    locations = cells_body(columns = "avg_improvement")
  ) |>
  tab_options(
    table.font.size = px(12),  # Adjust font size for readability
    heading.align = "left",    # Align header text to the left
    table.border.top.color = "black",  # Add a border at the top of the table
    table.border.bottom.color = "black"  # Add a border at the bottom of the table
  ) |>
  tab_source_note(
    source_note = md("**By:** Sam Benning  |  **Source:** [Basketball Reference](https://www.basketball-reference.com/)")
  )
```

I find it valuable to see the full picture of how each player has stacked up for their respective teams. A chart that features individual development will dive deeper into why a team stands where they are growth-wise.

```{r}
# Summarize total improvement for sorting
team_totals <- development_by_position %>%
  group_by(Team) %>%
  summarize(total_improvement = sum(improvement))

# Reorder teams by total improvement
development_by_position <- development_by_position %>% filter(Team != "2TM", Team != "3TM") |>
  mutate(Team = factor(Team, levels = team_totals$Team[order(team_totals$total_improvement)]))

# Create the plot
totalimprovement <- ggplot(development_by_position, aes(
  x = Team, 
  y = improvement, 
  fill = Pos, 
  text = paste0(Player, ", Improved PER by ", improvement, " (", years_with_team, " yrs)")
)) +
  geom_bar(stat = "identity", color = "gray90", size = 0.2, position = "stack") +  # Add border for separation
  coord_flip() +
  labs(
    title = "Player Improvement by Position (Using PER Growth as a Measure)",
    subtitle = "",
    x = "Team",
    y = "PER Growth While With Team"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 10),  # Adjust text size for readability
    panel.grid.major.y = element_line(color = "gray80")  # Optional grid for better alignment
  ) +
  guides(fill = guide_legend(title = "Position"))

### MAKING MY OWN FACET WRAP - R was being annoying
pg_dev <- development_by_position |> filter(Pos == "PG")
sg_dev <- development_by_position |> filter(Pos == "SG")
sf_dev <- development_by_position |> filter(Pos == "SF")
pf_dev <- development_by_position |> filter(Pos == "PF")
c_dev <- development_by_position |> filter(Pos == "C")

# Helper function to create a plot for a specific position
create_position_plot <- function(data, position) {
  ggplot(data, aes(
    x = reorder(Team, improvement, FUN = sum),  # Reorder teams by improvement
    y = improvement, 
    fill = Team, 
    text = paste0(Player, ", Improved PER by ", improvement, " (", years_with_team, " yrs)")
  )) +
    geom_bar(stat = "identity", color = "gray90", size = 0.25, position = "stack") +
    coord_flip() +
    labs(
      title = paste(position),
      x = "Team",
      y = "PER Growth While With Team (Individual players included)"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",  # Hide legend for individual plots
      axis.text.y = element_text(size = 10),  # Adjust for better readability
      panel.grid.major.y = element_line(color = "gray80")
    )
}

# Create individual plots for each position
pg_plot <- create_position_plot(pg_dev, "Point Guard")
sg_plot <- create_position_plot(sg_dev, "Shooting Guard")
sf_plot <- create_position_plot(sf_dev, "Small Forward")
pf_plot <- create_position_plot(pf_dev, "Power Forward")
c_plot <- create_position_plot(c_dev, "Center")
```

```{r}
#Plotting full graph, and each individual positional graph together
ggplotly(totalimprovement, tooltip = "text")
ggplotly(pg_plot, tooltip = "text")
ggplotly(sg_plot, tooltip = "text")
ggplotly(sf_plot, tooltip = "text")
ggplotly(pf_plot, tooltip = "text")
ggplotly(c_plot, tooltip = "text")
```

## Individual Player Development

There could be players who happen to just develop on their own - we want to give players credit for their own hard work, too. I next will look into players who have been on two or more teams in their career and will visualize who has been the best at improving themselves regardless of the team that they are on.

```{r}
#First, mutate to count the number of different teams a player has been with
all_data <- all_data %>%
  group_by(Player) %>%
  mutate(
    num_teams = n_distinct(Team, na.rm = TRUE)  # Count unique teams for each player
  ) %>%
  ungroup() |>
filter(!Team %in% c("2TM", "3TM", "4TM", "5TM")) |> #remove the variables that show both teams combined stats 
filter(!Year %in% 2025) #helps filter out players who might be doing well this year (small sample size)

#Filter for players who have played with \>1 team and record their stats as we have been doing
moved_players <- all_data |> filter(num_teams >=2) %>% filter(G > 20)|> group_by(Player)|>
  summarize(
    games = sum(G),
    min = sum(MP),
    seasons = max(Year) - mean(Draft_Yr),
    teams = mean(num_teams),
    rookie_PER = first(PER),
    peak_PER = max(PER, na.rm = TRUE),
    PER_growth = peak_PER - rookie_PER,
    ) |>
  arrange(desc(PER_growth))
```

Here are 8 of the top 10 players who developed very well on their own over the past 20 years regardless of team:

```{r}
# Filter data for a specific player
player_data <- all_data %>%
  filter(Player == "Domantas Sabonis" | Player == "James Harden" | Player == "Shai Gilgeous-Alexander" | Player == "Russell Westbrook" | Player == "Jimmy Butler" | Player == "Lou Williams" | Player == "Terry Rozier" | Player == "Kevin Durant") %>%
  mutate(yrspro = Year - Draft_Yr) |>
  group_by(Year) |>
arrange(yrspro)

# Create the line chart
ggplot(player_data, aes(x = yrspro, y = PER, color = Team)) +
  geom_line(size = 1) +  # Line for career trajectory
  geom_point(size = 3) +  # Points for individual seasons
  labs(
    title = "OKC is the Place to be",
    subtitle = "4 of the 8 current NBA players with the best career PER growth have played for OKC.",
    x = "Years Pro",
    y = "PER"
  ) +
  facet_wrap(~Player) +
  geom_text_repel(data=player_data, aes(x=yrspro, y=PER, label=Team), size = 2) +
  theme_minimal() +  # Minimal theme for clean visualization
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

I was looking for who had the best career trajectory and really stuck out the grind regardless of where they were. I was still left with my main takeaway coming from a team!

The OKC Thunder have clearly grown many stars through their organization. Obviously we are aware of the players who have come up through their organizations, but it is cool to visually see how players grow year-over-year as they progress with OKC. If we go back to the player development measures from before, they are in the top 10 of teams when it comes to those measures, too. The team knows what they are doing.

Another thing to credit these players who grew - (aside from the Thunder), a lot of growth occurred during years where a team wasn't contending. Rozier on Boston's crappy teams, then Charlotte...Lou Will on the Sixers...Butler with the Bulls (they got good as he got good)... and Sabonis with the Pacers. It seems like the ideal environment for the average player to make a name for themself is on a middling team. It makes me look forward to watching future drafts and trying to identify potential future stars.

If I were to keep going in this project, I would want to look at teams front offices as a whole and measure how good they are, using things like drafting, trading, contract handling, winning, etc. The research will continue as far as I can take it.
